Data Migration under a Changing Schema in Ampersand

### 2023-10-17 2022Migration/articleMigrationFACS.tex
making corrections in the migration script...

\section{Generating a Migration Script}
   A migration script takes care of two things: copying data and ensuring that the migration system is a proper information system.
   This section presents a general approach for generating a migration script.
   Let $\overleftarrow{\infsys}$ be the existing system, and let $\overrightarrow{\infsys}$ be the desired system.
   Consequently, we define $\pair{\overleftarrow{\dataset}}{\overleftarrow{\schema}}=\overleftarrow{\infsys}$ and $\pair{\overrightarrow{\dataset}}{\overrightarrow{\schema}}=\overrightarrow{\infsys}$ as the data set and schema of the existing and desired system, respectively.
   We also define $\quadruple{\overleftarrow{\concepts}}{\overleftarrow{\rels}}{\overleftarrow{\rules}}{\overleftarrow{\transactions}}=\overleftarrow{\schema}$
   and $\quadruple{\overrightarrow{\concepts}}{\overrightarrow{\rels}}{\overrightarrow{\rules}}{\overrightarrow{\transactions}}=\overrightarrow{\schema}$.

   Then we define the migration schema $\schema_\migrsys$ as follows:

   Let $\pair{\dataset}{\schema}$ be the existing system, and let ${\schema'}$ be the schema of the desired system.
   Then we define the migration systems $\migrsys$ and the events as follows:
\begin{enumerate}
\item We take a disjoint union of the relations, defining $\overleftarrow~$ and $\overrightarrow~$ as (overloaded) notation to generate new names, keeping the data set of the old system and the rules of the new system:
   \begin{align}
   \overleftarrow{\declare{n}{x}{y}}&=\declare{(\id{old}.n)}{x}{y}&\overleftarrow{X} &= \{\overleftarrow{x}\mid x\in X\}\\
   \overrightarrow{\declare{n}{x}{y}}&=\declare{(\id{new}.n)}{x}{y}& \overrightarrow{X} &= \{\overrightarrow{x}\mid x\in X\}\\
   \triples_\migrsys &= \{\triple{a}{\overleftarrow{r}}{b} \mid \triple{a}{r}{b}\in\triples_\dataset\}&
   \instance_\migrsys &= \instance_\dataset\\
   \rels_\theenumi &= \overleftarrow\rels_{\schema} \cup \overrightarrow\rels_{\schema'}
   %&\rules_\theenumi &= \rules_\schema
   \end{align}
\item We create transactions to copy equal relations:
   \begin{align}
   \widetilde{\declare{n}{x}{y}} &= \declare{(\id{copied}.n)}{x}{y}\\
   \rels_\theenumi &= \{\widetilde{r}\mid r \in \rels_\schema \cap \rels_{\schema'}\}\\
   \transactions_\theenumi &= \{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r} - \widetilde{r}} \mid \widetilde{r} \in \rels_2\} \cup \{\widetilde{r}\mapsfrom \popF{\overrightarrow{r}} \mid \widetilde{r} \in \rels_2\}
   \end{align}
\item For each rule, we generate two helper relations, {\tt viols} and {\tt fixed}.
   \begin{align}
   \overleftarrow{\dataset'} &= \pair{\{\triple{x}{{r}}{y}\mid \triple{x}{\overleftarrow{r}}{y} \in \triples_{\dataset'}\}}{\instance_{\dataset'}}\\
   \overrightarrow{\dataset'} &= \pair{\{\triple{a}{{r}}{b} \mid \triple{a}{\overrightarrow{r}}{b}\in\triples_\dataset\}}{\instance_{\dataset'}} \\
   \rels_\theenumi &= \{{\tt viols}_u\mid u \in \rules_{\schema'}\} \cup \{{\tt fixed}_u\mid u \in \rules_{\schema'}\}\\
   \rules_\migrsys &= \{\lambda {\dataset'}.~ \viol{u}{\overrightarrow{\dataset'}} - \pop{{\tt viols}_u - {\tt fixed}_u}{\dataset'} \mid u\in\rules_{\schema'}\}\label{eqn:migrsysrules}\\
   \transactions_{\theenumi} &=\{{\tt viols}_u \mapsfrom \lambda \dataset'.~ \viol{u}{\overleftarrow{\dataset'}}\} \cup
    \{{\tt fixed}_u \mapsfrom \lambda \dataset'.~ {\tt viols}_u - \viol{u}{\overrightarrow{\dataset'}}\}\label{eqn:enforceForRules}
   \end{align}
\item Combining the above into a single migration system:
   \begin{align}
   \concepts_\migrsys &= \concepts_\dataset \cup  \concepts_{\dataset'}\\
   \rels_\migrsys &= \rels_1\cup\rels_2\cup \rels_3\\
   %\rules_\migrsys &= \rules_1\cup \rules_3\\
   \transactions_\migrsys &= \transactions_1\cup\transactions_2\cup\transactions_3
   \end{align}
\end{enumerate}

\subsection{Validation}
   We first observe that $\migrsys = \pair{\pair{\triples_\migrsys}{\instance_\migrsys}}{\quadruple{\concepts_\migrsys}{\rels_\migrsys}{\rules_\migrsys}{\transactions_\migrsys}}$ is a system after application of the transactions:
   the only rules are those introduced in equation~\ref{eqn:migrsysrules}, and transactions from equation~\ref{eqn:enforceForRules} ensure that $\popF{{\tt viols}_u} = u$ initially, while $\popF{{\tt fixed}_u}$ is empty.
   Hence, the MoT can happen after the transactions have been applied, giving us $\migrsys'$.
   The event $\migrsys \xrightarrow[\rels_2 \cup \rels_3]{\overrightarrow\rels_{\schema'}}\migrsys'$ represents the MoT.

>>>>

\section{Generating a Migration Script}
   A migration script takes care of two things: copying data and ensuring that the migration system is a proper information system.
\subsection{Generating a migration script}
   In the migration system, we need to refer to the items (concepts, relations, and constraints) of both the existing system and the desired system.
   For this purpose, we relabel those items.
   We use a left arrow to denote relabeling by prefixing the name of the item with ``old.'', and a right arrow to denote prefixing with ``new.''.
\begin{equation}
   \begin{array}[m]{rcl}
      \overleftarrow{\pair{\dataset}{\schema}}&=&\pair{\overleftarrow{\dataset}}{\overleftarrow{\schema}}\\
      \overleftarrow{\pair{\triples}{\instance}}&=&\pair{\overleftarrow{\triples}}{\instance}\\
      \overleftarrow{\quadruple{\concepts}{\rels}{\rules}{\transactions}}&=&\quadruple{\concepts}{\overleftarrow{\rels}}{\overleftarrow{\rules}}{\overleftarrow{\transactions}}\\
      \overleftarrow{\triples}&=&\{\triple{a}{\overleftarrow{r}}{b}\mid\triple{a}{r}{b}\in\triples\}\\
      \overleftarrow{\declare{n}{A}{B}}&=&\declare{old.n}{A}{B}\\
      \overleftarrow{X}&=&\{\overleftarrow{x}\mid x\in X\}\\
      \viol{\overleftarrow{u}}{\overleftarrow{\dataset}}&=&\viol{u}{\dataset}\\
      \sign{\overleftarrow{u}}&=&\sign{u}\\
      \enfRel{\overleftarrow{u}}&=&\overleftarrow{\enfRel{u}}
   \end{array}
\end{equation}
   The notation $\overrightarrow{\infsys}$ is defined similarly, using prefix ``new.'' instead of ``old.''.

   Then we define the migration system $\migrsys$ as follows:
   Let $\pair{\dataset}{\schema}$ be the existing system, and let $\pair{\dataset'}{\schema'}$ be the desired system in its initial state.
\begin{enumerate}
\item We take a disjoint union of the data sets by relabeling relation names:
   \begin{align}
      \dataset_\migrsys={}&\overleftarrow{\dataset}\cup\overrightarrow{\dataset'}
   \end{align}
\item We create transactions to copy equal relations:
   \begin{align}
      \widetilde{\declare{n}{A}{B}}={}&\declare{(\id{copied}.n)}{A}{B}\\
      \begin{split}
         \transactions_1={}&\{\overrightarrow{r}\mapsfrom \popF{\overleftarrow{r} - \widetilde{r}} \mid r \in \rels_\schema \cap \rels_{\schema'}\}\\
         &\cup\{\widetilde{r}\mapsfrom \popF{\overrightarrow{r}} \mid r \in \rels_\schema \cap \rels_{\schema'}\}
      \end{split}
   \end{align}
\item For each new blocking invariant $u$, we generate two helper relations: ${\tt viols}_u$ and ${\tt fixed}_u$, and two corresponding transactions.
   \begin{align}
      \rels_\theenumi={}& \{{\tt viols}_u\mid u \in \rules_{\schema'}-\rules_{\schema}\} \cup \{{\tt fixed}_u\mid u \in \rules_{\schema'}-\rules_{\schema}\}\\
      \begin{split}
         \transactions_2={}& \{{\tt viols}_u \mapsfrom \violC{\overrightarrow{u}}\mid u \in \rules_{\schema'}-\rules_{\schema}\}\\
         &\cup \{{\tt fixed}_u \mapsfrom \lambda \dataset'.~ {\tt viols}_u - \viol{\overrightarrow{u}}{\dataset'}\mid u \in \rules_{\schema'}-\rules_{\schema}\}
      \end{split}\label{eqn:enforceForRules}
   \end{align}
\item Let us combine the above into a single migration schema, $\schema_\migrsys$:
   \begin{align}
      \concepts_\migrsys ={}& \concepts_\dataset \cup  \concepts_{\dataset'}\\
      \rels_\migrsys ={}& \overleftarrow\rels_{\schema} \cup \overrightarrow\rels_{\schema'} \cup \rels_3\\
      \rules_\migrsys={}& \emptyset\label{eqn:migrsysrules}\\
      \transactions_\migrsys ={}& \transactions_1\cup\transactions_2
   \end{align}
\end{enumerate}

\subsection{Validation}
   We first observe that $\migrsys = \pair{\dataset_\migrsys}{\quadruple{\concepts_\migrsys}{\rels_\migrsys}{\rules_\migrsys}{\transactions_\migrsys}}$ is a system after application of the transactions:
   the only rules are those introduced in equation~\ref{eqn:migrsysrules}, and transactions from equation~\ref{eqn:enforceForRules} ensure that $\popF{{\tt viols}_u} = u$ initially, while $\popF{{\tt fixed}_u}$ is empty.
   Hence, the MoT can happen after the transactions have been applied, giving us $\migrsys'$.
   The event $\migrsys \xrightarrow[\rels_2 \cup \rels_3]{\overrightarrow\rels_{\schema'}}\migrsys'$ represents the MoT.
